### 2025-10-06 thesis/design/index.tex
Improve design section layout

This chapter describes the design of \textsc{Natural}, the proposed NL2SQL system, 
addressing limitations and research gaps identified in the literature review.
The design follows a modular architecture that leverages both traditional computer
science algorithms and machine learning models to achieve robust, portable and
extensible natural language to SQL translation capabilities.

\subsection{Overview}

The \textsc{Natural} system implements a multi-stage pipeline that transforms natural
language questions into executable SQL queries through example selection,
schema subsetting, and iterative refinement. The proposed architecture consists
of five core components:

@@

The system processes queries through the following pipeline:

@@

This design largely builds upon in context learning concepts from DAIL-SQL \citep{DAIL-SQL}
and OmniSQL \citep{OmniSQL} but incorporates novel contributions in schema-aware example
selection by harnessing a graph-based structural similarity metric.

\subimport{}{initialization}
\subimport{}{functions}

>>>>

This chapter describes the design of \textsc{Natural}, our proposed NL2SQL system, 
addressing limitations and research gaps identified in the literature review.

\textsc{Natural} is architected as a pipeline that transforms natural
language questions into SQL queries using example selection ($\sigma$),
schema subsetting ($\phi$), and self refinement ($\rho$) and voting ($\nu$).
The proposed system consists of five components:

@@

The system processes queries using the following algorithm:

@@

This design largely builds upon few-shot learning concepts from DAIL-SQL \citep{DAIL-SQL}
and OmniSQL \citep{OmniSQL} but incorporates novel contributions in schema-aware example
selection by harnessing a graph-based structural similarity metric.

\subimport{}{initialization}
\subimport{}{functions}
\subimport{}{composition}

### 2025-10-06 thesis/design/initialization.tex
Improve design section layout

Given a database schema $s$, the corresponding graph $G_s = (V, E, \ell, w)$ is defined as:

\begin{enumerate}
    \item $V = V_t \cup V_c$ where $V_t$ represents table nodes and $V_c$ represents column nodes
    \item $E = E_{tc} \cup E_{fk} \cup E_{ref}$ where:
        \begin{itemize}
            \item $E_{tc}$ are table-column edges
            \item $E_{fk}$ are foreign key relationship edges between tables
            \item $E_{ref}$ are reference edges between foreign key columns
        \end{itemize}
    \item Each node $v \in V$ has a semantic label $\ell(v) \in \{1, 2, \ldots, 9\}$ where:
        \begin{itemize}
            \item $\ell(v) = 1$ for table nodes
            \item $\ell(v) = 2$ for generic column nodes
            \item $\ell(v) = 3$ for primary key columns
            \item $\ell(v) = 4$ for foreign key columns
            \item $\ell(v) = 5$ for text columns
            \item $\ell(v) = 6$ for numeric columns
            \item $\ell(v) = 7$ for datetime columns
            \item $\ell(v) = 8$ for boolean columns
            \item $\ell(v) = 9$ for other/unknown data types
        \end{itemize}
    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting structural importance:
        \begin{itemize}
            \item $w(e) = 1.0$ for foreign key relationships (highest importance)
            \item $w(e) = 0.9$ for column foreign key reference edges
            \item $w(e) = 0.5$ for table-column edges
        \end{itemize}
\end{enumerate}

We define the structural similarity of two databases as the topological distance
of the respective graphs. The graph representation captures essential schema
structure through semantic node labeling that prioritizes constraints over data
types: primary key columns receive label 3 regardless of their underlying data
type, foreign key columns receive label 4, and only then are remaining columns
categorized by data type (text=5, numeric=6, datetime=7, boolean=8, other=9).
This hierarchical labeling ensures that structural relationships take
precedence over semantic content, while omitting table names, column names and
domain terminology to achieve schema-agnostic comparison.

@@

For the database schemas above, the normalized schema (5 nodes, 5 edges) and
denormalized schema (3 nodes, 2 edges) show significant topological differences.
The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)
and the loss of relational structure (elimination of foreign key relationships),
resulting in a substantial distance value reflecting their structural dissimilarity
despite representing equivalent logical data with $D^{f}_{W}(G_{norm}, G_{denorm}) \approx 0.78$.

>>>>

Given a database schema $s$ we define the corresponding graph $G_s = (V, E, \ell, w)$ as:

\begin{enumerate}
    \item $V = V_t \cup V_c$ where $V_t$ represents table nodes and $V_c$ represents column nodes
    \item $E = E_{tc} \cup E_{fk} \cup E_{ref}$ where:
        \begin{itemize}
            \item $E_{tc}$ are table-column edges
            \item $E_{fk}$ are foreign key relationship edges between tables
            \item $E_{ref}$ are reference edges between foreign key columns
        \end{itemize}
    \item Each node $v \in V$ has a semantic label $\ell(v) \in \{1, 2, \ldots, 9\}$ where:
        \begin{itemize}
            \item $\ell(v) = 1$ for table nodes
            \item $\ell(v) = 2$ for generic column nodes
            \item $\ell(v) = 3$ for primary key columns
            \item $\ell(v) = 4$ for foreign key columns
            \item $\ell(v) = 5$ for text columns
            \item $\ell(v) = 6$ for numeric columns
            \item $\ell(v) = 7$ for datetime columns
            \item $\ell(v) = 8$ for boolean columns
            \item $\ell(v) = 9$ for other/unknown data types
        \end{itemize}
    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting its structural importance:
        \begin{itemize}
            \item $w(e) = 1.0$ for foreign key relationships (highest importance)
            \item $w(e) = 0.9$ for column foreign key reference edges
            \item $w(e) = 0.5$ for table-column edges
        \end{itemize}
\end{enumerate}

We define the structural similarity of two databases as the topological distance
of the respective graphs. The graph representation captures essential schema
structure through semantic node labeling that prioritizes constraints over data
types: primary key columns receive label 3 regardless of their underlying data
type, foreign key columns receive label 4, and only then are remaining columns
categorized by data type (text=5, numeric=6, datetime=7, boolean=8, other=9).
This hierarchical labeling ensures that structural relationships take
precedence over exact content, while omitting table names, column names and
domain terminology to achieve schema-agnostic comparison.

@@

For the database schemas above, the graph representation of the normalized
schema $G_{norm}$ (see figure~\ref{figure:normalized-graph-representation})
with 5 nodes and 5 edges and denormalized schema $G_{denorm}$ (see figure~\ref{figure:denormalized-graph-representation})
with 3 nodes and 2 edges show significant topological differences.
The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)
and the loss of relational structure (elimination of foreign key relationships),
resulting in a substantial distance value reflecting their structural dissimilarity
despite representing equivalent logical data with $D^{f}_{W}(G_{norm}, G_{denorm}) \approx 0.78$.
